---
title: "The Effects of pH on the in-vitro Growth of Four Fungal Pathogens of Processing Tomato"
output: github_document
author: "Vikesh Ajith"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300, fig.width = 7)
```

## Introduction

### Background

This notebook revisits the data from my first experiment in my 2019 Honours
thesis, "The effects of soil pH on infection of tomato by **Fusarium oxysporum**
(crown and root rot) and **Pythium spp.**". It is an informal analysis meant to
present how I would analyse this data today given my knowledge of programming
(The original analyses were done in Minitab).

Processing tomato farmers in Echuca-Moama (NSW/VIC) noticed that yields were
declining due to soilborne disease in fields where sub-surface drip irrigation
tubes had remained for a long period of time. This has been associated with
increased acidity in the soil surrounding the tubes. One possible hypothesis is
that the acidity promotes the growth of of soilborne pathogens, worsening
disease.

### Experimental Aims

The aim of this experiment was to investigate the effects of pH on the growth
rates of four known fungal pathogens of processing tomato (Fusarium oxysporum,
Pythium irregulare, Pythium catenulatum, and Pythium recalcitrans) in an
in-vitro medium (Potato Dextrose Agar)

### Methodology

Petri plates of PDA were prepared at 6 pH levels (4, 5, 5.5, 6, 7, 8) using HCl
and NaOH. These were inoculated with a 0.4mm plug of fungal culture, and culture
diameter was measured every 24 hours until the entire plate was colonised (up to
7 days).

Growth rates were calculated using the following formula: 
`(mean_width_cm -0.4)/day`. The growth rates used in the analysis were selected 
from the last day of growth common to all cultures within the species (This was 
to account for possible bias as some replicates took one day longer to fully 
colonise the plate).

## Data Cleaning

### Setup

```{r, message=FALSE}
library(tidyverse)
library(tidyxl)
library(unpivotr)
library(ggpubr)

library(hrbrthemes)
library(showtext)
library(scales)

font_add("IBMPlexSans", regular = "IBMPlexSans-Regular.ttf")
font_add("IBMPlexSans-Bold", regular = "IBMPlexSans-Bold.ttf")
font_add("IBMPlexSans-Medium", regular = "IBMPlexSans-Medium.ttf")
showtext_auto()
theme_set(theme_ipsum_ps(grid = "Yy", axis = T, ticks = F))
```

### Load Raw Data

```{r}
path_3 = "data/raw/2_invitro_ph/20190X00_ph_response_3.xlsx"
path_4 = "data/raw/2_invitro_ph/20190919_ph_response_4.xlsx"
path_5 = "data/raw/2_invitro_ph/20190919_ph_response_5.xlsx"
path_6 = "data/raw/2_invitro_ph/20190925_ph_response_6.xlsx"

# Extract sheets
pyth_cells_3 = lapply(1:5, xlsx_cells, path = path_3)
pyth_cells_4 = lapply(1:5, xlsx_cells, path = path_4)
fusa_cells_5 = xlsx_cells(path_5)
pyth_cells_6 = xlsx_cells(path_6)

# Initialise df lists for later
pyth_tidy_3 = list()
pyth_tidy_4 = list()
```

### Tidy Data

Normally, I would try to avoid repetition and use a function, however all the 
datasets vary slightly so it was easier to copy+paste than write a function 
that could deal with all the format variations.

```{r}
# Loop through sheets and tidy
for (cells in pyth_cells_3) {
    tidy = cells %>%
        filter(!is_blank & row <= 20 & col <= 25) %>%
        behead("up-left", "day") %>%
        behead("left", "comment") %>%
        behead("left", "species") %>%
        behead("left", "ph") %>%
        behead("left", "replicate") %>%
        behead("up", "dimension") %>%
        filter(dimension != "Mean") %>%
        rename(length_cm = numeric, comment = comment.header) %>%
        mutate(dimension = recode(dimension, W = "x", L = "y")) %>%
        add_column(trial = 3, start_date = NA) %>%
        pivot_wider(
            id_cols = c(start_date, trial, species, ph, replicate, day, comment),
            names_from = dimension,
            values_from = length_cm,
            names_sort = TRUE
        )
    pyth_tidy_3 = append(pyth_tidy_3, list(tidy))
}

pyth_tidy_3 = bind_rows(pyth_tidy_3)
```

```{r}
# Loop through sheets and tidy
for (cells in pyth_cells_4) {
    tidy = cells %>%
        filter(!is_blank) %>%
        behead("up-left", "day") %>%
        behead("left", "comment") %>%
        behead("left", "species") %>%
        behead("left", "ph") %>%
        behead("left", "replicate") %>%
        behead("up", "dimension") %>%
        filter(dimension != "Mean") %>%
        rename(length_cm = numeric, comment = comment.header) %>%
        mutate(dimension = recode(dimension, W = "x", L = "y")) %>%
        add_column(trial = 4, start_date = "2019-09-19") %>%
        pivot_wider(
            id_cols = c(start_date, trial, species, ph, replicate, day, comment),
            names_from = dimension,
            values_from = length_cm,
            names_sort = TRUE
        )
    pyth_tidy_4 = append(pyth_tidy_4, list(tidy))
}

pyth_tidy_4 = bind_rows(pyth_tidy_4)
```

```{r}
fusa_tidy_5 =
    fusa_cells_5 %>%
    filter(!is_blank) %>%
    behead("up-left", "day") %>%
    behead("left", "comment") %>%
    behead("left", "species") %>%
    behead("left", "ph") %>%
    behead("left", "replicate") %>%
    behead("up", "dimension") %>%
    filter(dimension != "Mean") %>%
    rename(length_cm = numeric, comment = comment.header) %>%
    mutate(dimension = recode(dimension, W = "x", L = "y")) %>%
    add_column(trial = 5, start_date = "2019-09-19") %>%
    pivot_wider(
        id_cols = c(start_date, trial, species, ph, replicate, day, comment),
        names_from = dimension,
        values_from = length_cm,
        names_sort = TRUE
    )
```

```{r}
pyth_tidy_6 =
    pyth_cells_6 %>%
    filter(!is_blank) %>%
    behead("up-left", "day") %>%
    behead("left", "comment") %>%
    behead("left", "species") %>%
    behead("left", "ph") %>%
    behead("left", "replicate") %>%
    behead("up", "dimension") %>%
    filter(dimension != "Mean") %>%
    rename(length_cm = numeric, comment = comment.header) %>%
    mutate(dimension = recode(dimension, W = "x", L = "y")) %>%
    add_column(trial = 5, start_date = "2019-09-25") %>%
    pivot_wider(
        id_cols = c(start_date, trial, species, ph, replicate, day, comment),
        names_from = dimension,
        values_from = length_cm,
        names_sort = TRUE
    )
```

### Clean Data

Convert `start_date` to dates, `day` and `replicate` to numeric, standardise 
`species` name to form `genus_species`, strip `pH` from `ph`. Bind all datasets
into one and save.

```{r}
dirty_datasets = list(
    pyth_tidy_3,
    pyth_tidy_4,
    fusa_tidy_5,
    pyth_tidy_6
    )

clean_datasets = list()

for (dataset in dirty_datasets) {
    clean_dataset =
        dataset %>%
        mutate(
            start_date = as.Date(start_date),
            day = as.numeric(str_extract(day, r"(\d+)")),
            replicate = as.numeric(replicate),
            species = str_replace_all(species,
                                      c(
                                          r"(P\. )" = "pythium_",
                                          r"(\. )" = "_",
                                          r"( )" = "_"
                                      )) %>% tolower(),
            ph = str_replace_all(
                ph,
                c(
                    r"(5.5 \(Control\))" = "5.5",
                    r"(Control)" = "5.5",
                    r"(pH )" = ""
                )
            ) %>% as.numeric(),
            comment = str_replace(comment, r"([Cc]ontam.*)", "contaminated")
        ) %>%
        list()
    clean_datasets = append(clean_datasets, clean_dataset)
}

ph_resp = 
    # bind datasets
    clean_datasets %>%
    bind_rows() %>%
    # Create unique ID for each biological replicate
    group_by(trial, species, ph, replicate) %>%
    # Add mean width and growth rate
    mutate(
        mean_width_cm = ave(x, y) %>% round(2),
        growth_rate_cm_day = ((mean_width_cm - 0.4)/day) %>% round(2),
        plate_id = cur_group_id()
    ) %>%
    select(plate_id, everything())
```

### Clean Environment

```{r}
rm(list=setdiff(ls(), "ph_resp"))
```

## Exploratory Data Analysis

### Generate Formatted Data

Rather than formatting data labels and axes for every graph, it was more 
convenient to just format the data itself. Dead and contaminated cultures were
also filtered.

```{r}
ph_resp_format = 
    ph_resp %>%
    # Take out dead/contaminated samples and redundant control 2
    filter(!comment %in% c("dead", "contaminated") & ph != "control_2") %>%
    # Format column names for printing to avoid formatting for every graph
    rename_with(str_to_title) %>%
    rename(
    "Start Date" = "Start_date",
    "Mean Width (cm)" = "Mean_width_cm",
    "Growth Rate (cm/day)" = "Growth_rate_cm_day",
    "pH" = "Ph",
    "Plate ID" = "Plate_id"
    ) %>%
    # Format species names for printing
    mutate(
        Species = str_replace_all(
            Species, c(r"(pythium_)" = "P. ",r"(fusarium_)" = "F. ")
        )
    )
```

### Summarise Data

First, lets take a look at the number of biological replicates below. 10 were
planned for each group, but there were some contamination and losses, most
notably for **P. recalcitrans** which had a tendency to contaminate easily. For
that species, some groups had 11 biological replicates as more were prepared to
account for potential losses.

```{r, message=FALSE}
# Take a look at the number of biological replicates per treatment
ph_resp_format %>%
    group_by(Species, pH, Day) %>%
    filter(Day == 1) %>%
    summarise(n = n()) %>%
    pivot_wider(names_from = Species, values_from = n) %>%
    select(!Day)
```
Next, lets determine the final cut-off day for each species. We can see from the
table and bar graph below that all cultures were growing till day 7, 4, 2, and 4
for the species F. oxysporum, P. catenulatum, P. irregulare, and P.
recalcitrans, respectively.

```{r, message=FALSE}
ph_resp_format %>%
    group_by(Species, Day) %>%
    summarise(n = n()) %>%
    pivot_wider(names_from = Day, values_from = n)

ph_resp_format %>%
    group_by(Species, Day) %>%
    summarise(n = n()) %>%
    ggplot(mapping = aes(x = Day, y = n)) +
        geom_bar(stat = "identity") +
        facet_wrap(~Species)
```
### Get Cut-off Data

```{r}
ph_resp_cut =
    ph_resp_format %>%
        filter(
            Species == "F. oxysporum" & Day == 7 |
            Species == "P. catenulatum" & Day == 4 |
            Species == "P. irregulare" & Day == 2 |
            Species == "P. recalcitrans" & Day == 4
        )
```

### Define pH scale colors

```{r}
ph_scale_fill = scale_fill_stepsn(
        colours = c(
            "red",
            "orange",
            "yellow",
            "greenyellow",
            "green4",
            "turquoise"
            ),
        n.breaks = 7
        )

ph_scale_color = scale_color_stepsn(
        colours = c(
            "red",
            "orange",
            "yellow",
            "greenyellow",
            "green4",
            "turquoise"
            ),
        n.breaks = 7
        )
```

### Bar Confidence Interval Plot

```{r}
ph_resp_cut %>%
    desc_statby(
        measure.var = "Growth Rate (cm/day)",
        grps = c("Species", "pH")
    ) %>%
    rename(`Mean Growth Rate (cm/day)` = mean) %>%
    ggplot(aes(
            x = pH,
            y = `Mean Growth Rate (cm/day)`,
            fill = pH
            )) +
    geom_bar(
        position = "dodge",
        stat = "identity"
        ) +
    geom_errorbar(
        aes(
            ymin = `Mean Growth Rate (cm/day)` - ci, 
            ymax = `Mean Growth Rate (cm/day)` + ci
            ),
        color = "gray65"
        ) +
    facet_wrap(~Species) +
    ph_scale_fill
```

```{r}
ph_resp_format %>%
    ggplot(
        aes(
            x = Day,
            y = `Mean Width (cm)`,
            color = pH,
            group = pH
            ),
        size = 1
        ) +
        geom_jitter(height = 0, width = 0.1) +
        geom_smooth(method = "lm", linetype = "solid", size = 0.3, se = F) +
        ph_scale_color +
        facet_wrap(~Species, nrow = 2)
```
```{r}
ph_resp_cut %>%
    ggplot(aes(x = pH, y = `Growth Rate (cm/day)`, group = Species), size = 1) +
    geom_jitter(height = 0, width = 0.1) +
    geom_smooth(linetype = "solid", size = 0.3) +
    ph_scale_color
```

```{r}
ph_resp_format %>%
    group_by(Species, pH, Day) %>%
    summarise(n()) %>%
    filter(Day == 1)

# ph_resp_format %>%
#     group_by(Trial, Species, pH, Replicate) %>%
#     filter(Day == max(Day)) %>%

ph_resp_format %>%
    group_by(Day) %>%
    View()

    # ggplot(
    #     mapping = aes(
    #         x = pH,
    #         y = `Growth Rate (cm/day)`,
    #         color = Species,
    #         group = Species,
    #         label = `Plate ID`
    #         )
    #     ) +
    # geom_jitter(height = 0, width = 0.1) +
    # geom_smooth() +
    # geom_label()
```

```{r}
ph_resp_format %>%
    ggqqplot(x = "Growth Rate (cm/day)") +
    facet_wrap(~Species)

# model = lm(
#     data = ph_resp_format,
#     formula = `Mean Width (cm)` ~ Day + pH + Species + Replicate + Trial
#     )
# 
# summary(model)
# anova(model, test = "F")
```